You have more money in your wallet. This does not mean you will spend extra on food

### 2024-04-16 paper/paper.qmd
updated parquet

---
title: "You have more money in your wallet. This does not mean you will spend extra on food"
subtitle: "Examining the relationship between disposable income and food expenditure for American consumers"
author: Ahnaf Alam
thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure"
date: today
date-format: long
classoption: abstract
abstract: "We examined the relationship between changes in food spending and increases in disposable income. Utilizing quarterly country-level data from the Federal Reserve Economic Data (FRED) spanning a 15-year period from 2007 to 2022, we employed both a simple linear regression model and a multiple regression model. Our analysis revealed a economically insignificant increase of $20 million in food spending for every billion-dollar rise in disposable income. From a policy perspective, our findings indicate that solely targeting disposable income is unlikely to substantially impact food spending behaviors. Therefore, policies designed to boost consumer spending should explore alternative approaches."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
margin-left: 25.4mm
margin-right: 25.4mm
---

@@

In @sec-data, we discuss the data we used for this project, including information of the software and packages used. This is followed by a explanation of the models that we used, which can be found in @sec-model. In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income. We finish the project with weaknesses and potential areas of improvement in the future.

@@

@@

We create this project using statistical software, `R` [@citeR]. For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2]. The data used in this paper comes from `fredr` [@fredr] package. We further used `rstanarm` [@rstanarm] for modelling. Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes.

@@

We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels. Utilizing the `fredr` package, we integrated six distinct datasets into one dataframe. These datatsets were:

@@

All datasets underwent seasonal adjustment and were indexed to 2017 dollars. Seasonal adjustments mitigate the influence of periodic fluctuations, such as strikes or unusual weather events, which can obscure genuine economic trends. By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods. Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames.

@@

FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S. Census Bureau, Bureau of Labor Statistics, and U.S. Bureau of Economic Analysis (BEA), among others. These datasets are aggregated at the country level. We use national-level data because it provides a broader perspective on general economic trends in the U.S. All five datasets we have used originate from the BEA. The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies. Upon receiving these datasets, BEA makes adjustments to the raw survey data. For instance, they fill in missing gaps in the datasets. For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from nonmerchant sellers. Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.

@@

@@

table1 %>%
  kable("latex", booktabs = TRUE) %>%
  column_spec(1, width = "2cm") %>%
  column_spec(2, width = "14cm") %>%
  kable_styling(latex_options = "hold_position", font_size = 9, position = "center")

@@

figure <- data_long |>
  ggplot( aes(x = date, y = value, fill = variable)) +
  geom_area() +
  labs(title = "Contribution of Expenditure Categories Over Time",
       x = "Date",
       y = "Expenditure",
       fill = "Categories") +
  theme_minimal() +
  scale_fill_manual(values = col, labels = newnames, name = "Categories")
print(figure)

@@

formatted_data <- data
formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) sprintf("%.2f", x))

formatted_data[1:12,] |> 
  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
                      "Healthcare", "Services"),
        booktabs = TRUE,
        linesep = "",
        align = "c",
        format.args = list(big.mark = ",")
        ) |>
  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
  kable_styling(font_size = 9, latex_options = "scale_down")  |>
  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the of the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
```

data |>
  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  theme_classic() + 
  labs(
    x = "Real increase in disposable income",
    y = "Real increase in food expenditure"
  )
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-3
#| tbl-cap: Summary results for both models

@@

@@

Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit. Notably, the root mean square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.

@@

@@

Current research suggests that higher income tends to result in increased expenditure on durable goods, surpassing that of other consumable products. Consumers often allocate the majority of their additional income towards items that provide utility over an extended period rather than those consumed in a single instance. Using a linear probability function, @lee1964income showed that consumers typically increase their spending on durable goods in response to income growth and resist reducing it even when income decreases. This pattern was evident during the COVID-19 pandemic, where heightened disposable income led to above-average spending on durable goods such as furniture, appliances, motor vehicles, and recreational items [@tauber2021why]. Therefore, it is plausible that food expenditure remains relatively stable compared to expenditures on durable goods, which may fluctuate more significantly in response to changes in income.

>>>>

---
title: "You have more money in your wallet. This does not mean you will spend extra on food"
subtitle: "Examining the relationship between disposable income and food expenditure for American consumers from 2007-2022"
author: Ahnaf Alam
thanks: "Code and data are available at https://github.com/AhnafAlam1/income_and_expenditure"
date: today
date-format: long
classoption: abstract
abstract: "We examined the relationship between changes in food spending and increases in disposable income. Utilizing quarterly country-level data from the Federal Reserve Economic Data (FRED) spanning a 15-year period from 2007 to 2022, we employed both a simple linear regression model and a multiple regression model. Our analysis revealed a economically insignificant increase of $20 million in food spending for every billion-dollar rise in disposable income. From a policy perspective, our findings indicate that solely targeting disposable income is unlikely to substantially impact food spending behaviors. Therefore, policies designed to boost consumer spending should explore alternative approaches."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
margin-left: 25.4mm
margin-right: 25.4mm
---

@@

In @sec-data, we discuss the data we used for this project, including information of the software and packages used. This is followed any a explanation of the models that we used, which can be found in @sec-model. In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income. We finish the project with weaknesses and potential areas of improvement in the future.

In @sec-data, we discuss the data we used for this project, including information of the software and packages used. This is followed by a explanation of the models that we used, which can be found in @sec-model. In @sec-res, we present the results of the models and @sec-dis, we interpret these results, and consider how savings, consumer preference and elasticies play into this discussion on food expenditure and income. We finish the project with weaknesses and potential areas of improvement in the future. Datasheet and model checks can be found under [Appendix -@sec-model-details].

@@

table1 %>%
  kable("latex", booktabs = TRUE) %>%
  column_spec(1, width = "2cm") %>%
  column_spec(2, width = "14cm") %>%
  kable_styling(latex_options = "hold_position", font_size = 9, position = "center")
```

@@

We create this project using statistical software, `R` [@citeR]. For cleaning and re-purposing the data, we used `tidyverse` [@tidy] package and graphs, we relied on `ggplot2` [@ggplot2]. The data used in this paper comes from `fredr` [@fredr] package. We further used `rstanarm` [@rstanarm] and `modelsummary` [@ms] for modelling. Lastly, we used `kableExtra` [@kable] and `viridis` [@viridis] for aesthetics purposes. For editing texts, we used `rspell` library [@rspell].

formatted_data <- data
formatted_data[columns_to_format] <- lapply(data[columns_to_format], function(x) formatC(x, format = "f", digits = 2, big.mark = ","))

formatted_data[1:12,] |> 
  kable(col.names = c("Date", "Durable goods", "Non-durable goods", "Food", "Disposable income",
                      "Healthcare", "Services"),
        booktabs = TRUE,
        linesep = "",
        align = "c",
        format = "latex") |> 
  column_spec(c(1, 2, 3, 4, 5, 6, 7), width = "6em") |> 
  kable_styling(font_size = 9, latex_options = "scale_down")  |>
  add_footnote(c("Note: The table displays data from 2007-2009 to illustrate the aggregated clean data. Including the entire dataset would be redundant and unnecessarily clutter the display."))
```

@@

We gathered our data from Federal Reserve Economic Data (FRED), an extensive online database housing a multitude of economic time series, both at the US national and international levels. Utilizing the `fredr` package, we integrated six distinct datasets into one data frame. These datasets were:

@@

All datasets underwent seasonal adjustment and were indexed to 2017 dollars. Seasonal adjustments mitigate the influence of periodic fluctuations, such as strikes or unusual weather events, which can obscure genuine economic trends. By referencing all data to the 2017 price level, we adjusted for inflation over time, facilitating accurate comparisons across various periods. Additionally, utilizing real economic data rather than nominal figures allowed for valid comparisons between different categories across time frames

@@

FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S. Census Bureau, Bureau of Labor Statistics, and U.S. Bureau of Economic Analysis (BEA), among others. These datasets are aggregated at the country level. We use national-level data because it provides a broader perspective on general economic trends in the U.S. All five datasets we have used originate from the BEA. The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies. Upon receiving these datasets, BEA makes adjustments to the raw survey data. For instance, they fill in missing gaps in the datasets. For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers. Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.

@@

All datasets underwent rigorous cleaning and preparation by both FRED and BEA before being made available on the FRED website. Consequently, we found no need for extensive cleaning on our part. The data was complete with no missing values, and both columns and rows were appropriately labeled and displayed. For better readability, we reorganized the columns and consolidated all five datasets into one dataframe, which we used for analysis.

@@

@@

figure <- data_long |>
  ggplot( aes(x = date, y = value, fill = variable)) +
  geom_area() +
  labs(title = "Contribution of Expenditure Categories Over Time",
       x = "Date",
       y = "Expenditure (in billions of 2017 USD)",
       fill = "Categories") +
  theme_minimal() +
  scale_fill_manual(values = col, labels = newnames, name = "Categories")
print(figure)

@@

@@

data |>
  ggplot(aes(x = income_expenditure, y = food_expenditure)) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  theme_classic() + 
  labs(
    x = "Real increase in disposable income",
    y = "Real increase in food expenditure"
  )
```

@@

Both models exhibit similar adjusted $R^2$ values, yet the second model slightly outperforms the first with a higher value (0.993 vs 0.990), indicating a superior fit. Notably, the root-mean-square error (RSME) for the second model is substantially lower than that of the first (10.36 vs 27.49), indicating enhanced predictive accuracy.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-3
#| tbl-cap: Summary results simple linear regression and multiple linear linear regression.

@@

Current research suggests that higher income tends to result in increased expenditure on durable goods, surpassing that of other consumable products. Consumers often allocate the majority of their additional income towards items that provide utility over an extended period, rather than those consumed in a single instance. Using a linear probability function, @lee1964income showed that consumers typically increase their spending on durable goods in response to income growth and resist reducing it even when income decreases. This pattern was evident during the COVID-19 pandemic, where heightened disposable income led to above-average spending on durable goods such as furniture, appliances, motor vehicles, and recreational items [@tauber2021why]. Therefore, it is plausible that food expenditure remains relatively stable compared to expenditures on durable goods, which may fluctuate more significantly in response to changes in income.
